---
title: "Day 11 - Homework Responses"
author: "YOUR NAME HERE"
format: gfm
---

## Preliminaries
```{r}
#| echo: FALSE
#| message: FALSE

# load up the packages
library(tidyverse)
library(sf)
library(here)   ### Helps with file management directories


# Read in the zipcode shapefile
zips <- st_read(dsn = here::here("day11","shapefiles","Zip_Codes"), layer = "Zip_Codes")

```

Link to google doc for instructions: [LINK](https://docs.google.com/document/d/1shtTHzs8y09XZd86nKNFVSGjtsa-s8XVoalSJ5kyDpk/edit?usp=sharing)


<div style="background-color: #f0f8ff; border-left: 5px solid #007acc; padding: 10px;">
## (20 points) Code of Ethics
I'm interested in understanding your process for completing all the questions in this assignment. Please tell me about the tools and resources—including any use of AI—that helped you with your work. Please remember to follow the outlined [Code of Ethics](https://docs.google.com/document/d/1lMvzPTGUAaMnH0KPibShQ1ualuo8uqICCexkh8zmsW8/edit?usp=sharing).  Because our understanding of how to work with AI tools is emerging, please err on the side of including too much detail rather than too little.


TYPE YOUR RESPONSE HERE

</div>




## Problem 1 (10 points)
Review [Chapter 18 in Modern Data Science with R](https://mdsr-book.github.io/mdsr2e/ch-spatial2.html).  Now that we have had a few days on shapefiles and maps, what interesting things did you learn about working with shapefiles?

A big thing that stood out to me was that shapefiles aren't just map making tools, but they are also data that can be directly involved in analysis, without being used as a map.  For example, you could find distances, areas, and lengths all directly from functions like st_distance() or st_area().  This is a different way of thinking about this data because I immediately associate mpas with strictly visual, but shapefiles act as a map in number form, and they are way more versatile.

Another things that was very interesting, and seemed useful, was spatial joins.  Instead of joining through columns (like usual), it is a join based on actual geographic location.  That's a completely new and different way of thinking about joing compared to what we've previously done in R.  With that being said, it is quite a powerful tool, datasets can be connected that don't actually have shared variables, they just share the same geographic space.  Overall, this read introduced shapefiles as more than just a static map, it showed that they are versatile building blocks for analysis.

## Problem 2 (10 points)
Read Chapters 1 and 2 and skim a few other chapters of interest from [Analyzing US Census Data: Methods Maps, and Models in R](https://walker-data.com/census-r/index.html). What did you learn about the US Census Data and how could you use that to level up your data science skills?

TYPE RESPONSE BELOW

## Problem 3 (20 points)
In our work in class, when we glimpsed the data frame `hennepin_zips`, more data than just population was included.  Select another variable and make an aggregate chlorpleth map of the zip code areas for Hennepin County. Use either `ggplot`, `mapview`, or `tmap`.

```{r}
#| echo: false
#| message: false

# load up the packages
library(zipcodeR)


# Read in the zipcodes in Hennepin County
hennepin_zips <- search_county("Hennepin","MN")


# Type your response here
```


## Problem 4 (20 points)
(10 points) The R package [tidycensus](https://walker-data.com/tidycensus/index.html) has information at the census tract level from the American Community Survey or the Decennial census. Read through the basic usage of this package: [LINK](https://walker-data.com/tidycensus/articles/basic-usage.html).

Discuss with your Hennepin County group the types of variables collected from the American Community Survey.  What variables might be of use in your project.  (One person can submit on behalf of your group. Since you are submitting individual homework, if you aren’t the person submitting, please tell me who is submitting in your response.)

I encourage you to start working with this package - notice that you need an API key, but it is easy to obtain.

TYPE RESPONSE BELOW
